$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-04 paper/paper.tex
on properties

\todo{intro}

\subsection{Framing}
\label{framing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Framing captures the idea that the well-typedness of a process is independent of its leftover resources.

\begin{nitheorem}
  \label{thm:framing}
  Let $P$ be a well-typed process $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$.
  Then, per \autoref{lm:types-op}, there exists some $\Delta$ such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Take arbitrary contexts $\Gamma_r$ and $\Xi_r$ such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\subsection{Subject Congruence}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{subject-congruence}

@@

\begin{proof}
  By induction on \textsc{Equals}.
  The base cases under $\constr{struct}$ (and their symmetric variants under $\constr{sym}$) are the nontrivial ones.
  For those, we proceed by induction on \textsc{StructCong}:
  \begin{itemize}
    \item
      The case $\constr{comp-assoc}$ is trivial: leftover typing is naturally associative.
    \item
      In the $\constr{comp-sym}$ case for a process $\comp{P}{Q}$, we need framing (\autoref{thm:framing}) to shift the output context of $P$ to the one of $Q$; and the input context of of $Q$ to the one of $P$.
      The point at which $Q$ and $P$ join in $\comp{Q}{P}$ will no longer be the same.
    \item
      The case $\constr{comp-end}$ is trivial: the typing rule for a process $\PO$ has the same input and output contexts.
    \item
      In the $\constr{scope-end}$ case, we show that the usage annotation of the newly created channel must be $\lz$, making the proof trivial.
      In the opposite direction, we instantiate the newly created channel to a type $\unit$ and a usage annotation $\lz$..
    \item
      In the $\constr{scope-ext}$ case for a process $\new \comp{P}{Q}$, we use \autoref{lm:types-unused} to show that $P$ preserves the usage annotations of the unused variable.
      We then use strengthening (\autoref{thm:strengthening}) to delete the unused variable from $P$, and compose the result with typing derivation on $Q$ wrapped in $\constr{chan}$.
      In the reverse direction, we weaken (\autoref{thm:weakening}) $P$ so that it accepts the additional variable.
      We then use \autoref{lm:lower-lift} to show that lowering and then lifting $P$ results in $P$.
    \item
      In the $\constr{scope-comm}$ case, we use swapping (\autoref{thm:swapping}).
      In the reverse direction, we use swapping and then \autoref{lm:swap-swap} to show that swapping two elements in $P$ twice leaves $P$ unchanged.
  \end{itemize}
\end{proof}

@@

\autoref{thm:subst-generalization} proves a generalized version of substitution, where the substitution $\subst{P}{j}{i}$ is on any variable $i$.
\autoref{thm:substitution} instantiates the generalized version to the concrete case where $i$ is $\constr{0}$, as required by subject reduction.

\begin{nitheorem}
  \label{thm:subst-generalization}
  Let process $P$ be well-typed such that $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$,
  Let $\ni_i m$ be a \textsc{VarRef} $\contains{\gamma}{\Gamma_i}{i}{t}{m}{\Gamma}$ for some $m$ and $\Gamma$.
  Let $\ni_i n$ be a \textsc{VarRef} $\contains{\gamma}{\Psi_i}{i}{t}{n}{\Psi}$ for some $m$ and $\Psi$.
  Let $\Gamma$ and $\Psi$ be related such that $\opctx{\Gamma}{\Delta}{\Psi}$ for some $\Delta$.
  Let $\Delta$ have a usage annotation $\lz$ at position $i$.
  Let $\ni_j m$ be a \textsc{VarRef} $\contains{\gamma}{\Gamma_j}{j}{t}{m}{\Gamma}$ for some $\Gamma_j$.
  Let $\ni_j n$ be a \textsc{VarRef} $\contains{\gamma}{\Psi_j}{j}{t}{n}{\Psi}$ for some $\Psi_j$.
  Then substituting $i$ with $j$ in $P$ will be well-typed such that $\types{\gamma}{\Gamma_j}{\subst{P}{j}{i}}{\Psi_j}$.
  Refer to \autoref{fig:subst} for a diagramatic representation.
\end{nitheorem}

@@

\begin{proof}
  By induction on the derivation $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  \begin{itemize}
    \item
      For constructor $\constr{end}$ we get $\Gamma_i \equiv \Psi_i$.
      From $\Delta_i \equiv \lz$ follows that $m \equiv n$.
      Therefore $\Gamma_j \equiv \Psi_j$ and $\constr{end}$ can be applied.

@@

\begin{proof}
  \hfill{}\\
  \begin{itemize}
    \item
    In the $\constr{comm}$ case we apply framing (\autoref{thm:framing}) to rearrange the assumptions in \autoref{fig:subject-reduction:a} into the proofs in \autoref{fig:subject-reduction:b}.
    We then apply substitution (\autoref{thm:substitution}) and strengthening (\autoref{thm:strengthening}).

>>>>

\todo{intro}
\todo{I wonder if we can remove the proofs everywhere and comment on them in the intro and where they can be found.}
\subsection{Framing}
\label{framing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Framing captures the idea that the well-typedness of a process is independent of its leftover resources.
\todo{Do not forward refer when it comes to results. Never refer to results in the appendix. Consider that, if the paper get publshed, it will get so without the appendix. It should be *completely* self contained}
\begin{nitheorem}
  \label{thm:framing}
  Let $P$ be a well-typed process $\types{\gamma}{\Gamma_l}{P}{\Xi_l}$.
  Then, per \autoref{lm:types-op}, there exists some $\Delta$ such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Take arbitrary contexts $\Gamma_r$ and $\Xi_r$ such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\todo{Maybe all previous subsections could be put into one "Auxiliary Results"}
\todo{Replace the proofs with an intuition on the theorem}
\subsection{Subject Congruence}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{subject-congruence}

@@

\todo{considering that Equals is in the appendix, you cannot do induction on something that is not defined in the main text. Bring the proof in the appendix. Maybe briefly comment on the proof here}
\begin{proof}
  By induction on \textsc{Equals}.
  The base cases under $\constr{struct}$ (and their symmetric variants under $\constr{sym}$) are the nontrivial ones.
  For those, we proceed by induction on \textsc{StructCong}:
  \begin{itemize}
    \item
      The case $\constr{comp-assoc}$ is trivial: leftover typing is naturally associative.
    \item
      In the $\constr{comp-sym}$ case for a process $\comp{P}{Q}$, we need framing (\autoref{thm:framing}) to shift the output context of $P$ to the one of $Q$; and the input context of of $Q$ to the one of $P$.
      The point at which $Q$ and $P$ join in $\comp{Q}{P}$ will no longer be the same.
    \item
      The case $\constr{comp-end}$ is trivial: the typing rule for a process $\PO$ has the same input and output contexts.
    \item
      In the $\constr{scope-end}$ case, we show that the usage annotation of the newly created channel must be $\lz$, making the proof trivial.
      In the opposite direction, we instantiate the newly created channel to a type $\unit$ and a usage annotation $\lz$..
    \item
      In the $\constr{scope-ext}$ case for a process $\new \comp{P}{Q}$, we use \autoref{lm:types-unused} to show that $P$ preserves the usage annotations of the unused variable.
      We then use strengthening (\autoref{thm:strengthening}) to delete the unused variable from $P$, and compose the result with typing derivation on $Q$ wrapped in $\constr{chan}$.
      In the reverse direction, we weaken (\autoref{thm:weakening}) $P$ so that it accepts the additional variable.
      We then use \autoref{lm:lower-lift} to show that lowering and then lifting $P$ results in $P$.
    \item
      In the $\constr{scope-comm}$ case, we use swapping (\autoref{thm:swapping}).
      In the reverse direction, we use swapping and then \autoref{lm:swap-swap} to show that swapping two elements in $P$ twice leaves $P$ unchanged.
  \end{itemize}
\end{proof}

@@

\autoref{thm:subst-generalization} proves a generalized version of substitution, where the substitution $\subst{P}{j}{i}$ is on any variable $i$.
\autoref{thm:substitution} instantiates the generalized version to the concrete case where $i$ is $\constr{0}$, as required by subject reduction.
\todo{Structure the theorem (in general the long ones) a bit more. Use itemize or similar. It's a blob of text.}
\begin{nitheorem}
  \label{thm:subst-generalization}
  Let process $P$ be well-typed such that $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$,
  Let $\ni_i m$ be a \textsc{VarRef} $\contains{\gamma}{\Gamma_i}{i}{t}{m}{\Gamma}$ for some $m$ and $\Gamma$.
  Let $\ni_i n$ be a \textsc{VarRef} $\contains{\gamma}{\Psi_i}{i}{t}{n}{\Psi}$ for some $m$ and $\Psi$.
  Let $\Gamma$ and $\Psi$ be related such that $\opctx{\Gamma}{\Delta}{\Psi}$ for some $\Delta$.
  Let $\Delta$ have a usage annotation $\lz$ at position $i$.
  Let $\ni_j m$ be a \textsc{VarRef} $\contains{\gamma}{\Gamma_j}{j}{t}{m}{\Gamma}$ for some $\Gamma_j$.
  Let $\ni_j n$ be a \textsc{VarRef} $\contains{\gamma}{\Psi_j}{j}{t}{n}{\Psi}$ for some $\Psi_j$.
  Then substituting $i$ with $j$ in $P$ will be well-typed such that $\types{\gamma}{\Gamma_j}{\subst{P}{j}{i}}{\Psi_j}$.
  Refer to \autoref{fig:subst} for a diagramatic representation.
\end{nitheorem}

@@

\todo{I don't think a reviewer from CONCUR most likely will nto get the proof. Appendix! Replace it with a proof sketch where you comment on the diagram as well.}
\begin{proof}
  By induction on the derivation $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  \begin{itemize}
    \item
      For constructor $\constr{end}$ we get $\Gamma_i \equiv \Psi_i$.
      From $\Delta_i \equiv \lz$ follows that $m \equiv n$.
      Therefore $\Gamma_j \equiv \Psi_j$ and $\constr{end}$ can be applied.

@@

\todo{maybe only focus on the first case, internal and external and say the rest is by induction blah bla and details can be found in appenix/or github}
\begin{proof}
  \hfill{}\\
  \begin{itemize}
    \item
    In the $\constr{comm}$ case we apply framing (\autoref{thm:framing}) to rearrange the assumptions in \autoref{fig:subject-reduction:a} into the proofs in \autoref{fig:subject-reduction:b}.
    We then apply substitution (\autoref{thm:substitution}) and strengthening (\autoref{thm:strengthening}).
